\documentclass[doctype=phd,lang=en,bibstyle=apa]{ulthesis}

\usepackage{mystyle}

% Cover content
\department {Departamento de Informática}
\title      {Semantic Similarity Across\\Biomedical Ontologies}
\programme  {Doutoramento em Informática}
\branch     {Especialidade Bioinformática}
\author     {João Diogo Silva Ferreira}
\advisor    {Prof. Dr. Francisco José Moreira Couto}
\date       {2016}

% References
%\addbibresource{content/references.bib}
\addbibresource{content/references-mendeley.bib}

% During typesetting we might be interested in one chapter at a time
% \includeonly{
%     frontmatter/dedicatory,
%     frontmatter/acknowledgements,
%     content/00-abstract-pt,
%     content/00-abstract-en,
%     content/01-introduction,
%     content/02-concepts,
%     content/03-state-of-the-art,
%     content/04-validation,
%     content/05-enhancements,
%     content/05a-interlude,
%     content/06-data-description,
%     content/07-multidomain,
%     content/08-technical,
%     content/09-conclusions,
%     content/appendix-a,
%     content/appendix-b,
% }


% Start document
\begin{document}

\setcounter{page}{-6}
\pagenumbering{gobble}

\coverpage

\include{frontmatter/dedicatory}
\include{frontmatter/acknowledgements}

% Start numbering pages with Roman numerals
\frontmatter

% Abstracts in several languages
\startabstract[pt]\include{content/00-abstract-pt}
\startabstract[en]\include{content/00-abstract-en}

% Lists in the front matter
\cleardoublepage
\tableofcontents*
{\def\mathversion#1{}\cleardoublepage\listoffigures}
{\def\textbf{}\cleardoublepage\listoftables}
\cleardoublepage\listoflistings

% \cleardoublepage\listoftodos

% Switch to Arabic numeration
\mainmatter

\part{Background}
\epigraph{Imagination is the beginning of creation. You imagine what you desire, you will what you \penalty-500 imagine and at last you create what you will.}{George Bernard Shaw}

\include{content/01-introduction}
\include{content/02-concepts}
\include{content/03-state-of-the-art}

\part{Contributions}
\epigraph{Scientists have become the  bearers of the torch of discovery in our quest for knowledge.}{Stephen Hawking}

\include{content/04-validation}
\include{content/05-enhancements}
\include{content/05a-interlude}
\include{content/06-data-description}
\include{content/07-multidomain}
\include{content/08-technical}

\part{Final remarks}
\epigraph{I think and think for months and years. Ninety-nine times, the conclusion is false. The hundredth time I am right.}{Albert Einstein}

\include{content/09-conclusions}

\part{Back Matter}
\epigraph{The moment a man sets his thoughts down on paper, he is in a sense writing for publication.}{Raymond Chandler}

\appendix
\include{content/appendix-a}
\include{content/appendix-b}

\small
\printbibliography

\end{document}
