\cleardoublepage
\pdfbookmark{Acknowledgements}{Acknowledgements}
\chapter*{Acknowledgements}

Foremost, I would like to express my gratitude to my advisor, Prof.~Dr~Francisco M.\ Couto, whose expertise and undeniably scientifically oriented mind added considerably to my PhD. I appreciate his vast horizons and skills in many scientific areas (\eg semantic similarity, biomedical ontologies, Bioinformatics in general, just to list the ones more closely related to my research efforts), and his assistance in the extensive periods where writing was the major undertaking of my work (particularly the several manuscripts we jointly submitted for publication, as well as this document). Most of all, I look up to his ``relaxed'' yet serious attitude towards research, which always positively surprised me. Sometimes, research feels like a herculean task that we need to carry out in order to see some results, but more often it looks like a small train ride where we simply appreciate the landscape of knowledge to which we ourselves contribute.

My PhD companions, which are no longer students but Doctors, also deserve my gratitude: in no particular order, I thank Ana Teixeira, Catia M. Machado, Catia Pesquita, Daniel Faria, Hugo Bastos and Tiago Grego, with whom I collaborated in some way or another, and whom I could always count on to tag along on a brainstorming session over coffee. A further acknowledgement goes to André Lamúrias, with whom I collaborated towards the end of my PhD.

I recognize that this research would not have been possible without the financial assistance of the Fundação para a Ciência e Tecnologia, and express my gratitude to that agency for funding my PhD scholarship SFRH\slash BD\slash 69345\slash 2010.

I would also like to thank my parents for the infinite support they have been providing throughout my entire life, from the very first breath I took. The excellent upbringing you provided was the exact amount of motivation I needed to pursue the career in science that I now so dearly esteem. Thank you for always being there for me, for always helping me succeed. I also want to thank my brothers, André and Simão, for no other motive than that you are always present in my life. You are the best.

In particular, I must acknowledge my wife and best friend, Sandra, who stood by me every step of the way, who had to endure my absence whenever I needed to attend some conference, and without whose love, encouragement and unconditional assistance I would not have been able to finish this work. The biggest of my smiles is yours.

\vfill

{\flushright
\textit{Lisboa}, January 6\textsuperscript{th}, 2016\break
João Ferreira
\par}
